---
title: "gene ontology"
output:
  rmdformats::html_clean:
    code_folding: show
    default_style: light
    toc_depth: 6
    fig_caption: yes
---
## Gene ontology enrichment using topGO
Question: are there any patterns in the types of genes lost (present on X, missing on Y?)
### Loading packages and data
```{r}
library("tidyverse")
library("httr")
library("Rgraphviz")
library("biomaRt")
library("topGO")
library("GO.db")
#genelist<-read_table("genelisthap1.txt")
### list of blast-confirmed yloss genes
genelist_blast<-read_table("genelisthap1_blast.txt")
### list of genes that pass the filter used in the denominator of missingness analysis
denomGenes<-read_delim("denomgenelist_salorth.txt") %>% 
  filter(!(grepl("\\|",hap1))) #%>% dplyr::select(hap1) 
### raw annotation file containing genes w annotated GO terms (not all genes have)  
annotation<-read_delim("NChap1_final.all.gff",comment = "#",col_names = F, delim = "\t")
### parsing to extract GO terms, IDs, etc.
GO_anno <- annotation %>% filter(X3 == "gene" ) %>% dplyr::select(X9) %>% 
  separate(X9,into=c("anno","GOterm"),sep ="Ontology_term=") %>%
  separate(GOterm, into=c("GOterm","notes"),sep="Note=") #%>%
  #filter()
### create a gene GOs mapping file - matches example in topGO. 
# Gene names and comma separated GO terms
GO_map<- GO_anno %>% filter(!is.na(GOterm)) %>%  
  mutate(ID = str_replace(anno, ";.+", "")) %>% mutate(ID = str_replace(ID, "ID=", "")) %>% 
  mutate(ID = str_replace(ID, "-RA", "")) %>% 
  mutate(GOterm = str_replace(GOterm, ";","")) %>%
  dplyr::select(ID,GOterm) %>% distinct(ID, GOterm, .keep_all = TRUE)
# same as above but only include the denominator set from the missinggenes R script
GO_map_denom<-GO_map %>% filter(ID %in% denomGenes$hap1) 
#only 5000? Not all genes have a go term and not all in GO_map have sal orth
## writing to .map file. readMappings() only accepts as an external file
write.table(GO_map_denom,"geneIDsGO_denom.map", row.names=F,col.names = F,quote = F)
```

### Read in gene:GOs mapping file
```{r}
geneID2GO<-readMappings(file="geneIDsGO_denom.map",sep=" ",IDsep=",") 

geneNames <- names(geneID2GO)
```
### filter interesting genes by X genes only
Get genes missing from Y on X only - for now. Opportunity to include autosomal missing genes later on.
```{r}
GO_anno_X <- annotation %>%
  filter(X1 == "X" ) %>%
  filter(X3 == "gene" )  %>% 
  separate(X9,into=c("anno","GOterm"),sep ="Ontology_term=") %>%
  separate(GOterm, into=c("GOterm","notes"),sep="Note=") %>%
  mutate(anno = str_replace(anno, ";.+", "")) %>% 
  mutate(anno = str_replace(anno, "ID=", "")) %>% dplyr::select(anno)
Xgenelist<-genelist_blast %>% filter(hap1 %in% GO_anno_X$anno)

```
### create GOdata object and interesting gene list

```{r}
MyInterestingGenes_X<-as.character(Xgenelist$hap1)
geneList <- factor(as.integer(geneNames %in% MyInterestingGenes_X))
names(geneList) <- geneNames
str(geneList)
GOdataBP_X <- new("topGOdata", ontology = c("BP"), allGenes = geneList,description = "X genes in total set",
  geneSel = MyInterestingGenes_X, annot = annFUN.gene2GO, gene2GO = geneID2GO, nodeSize = 10)
GOdataMF_X <- new("topGOdata", ontology = c("MF"), allGenes = geneList,
  geneSel = MyInterestingGenes_X, annot = annFUN.gene2GO, gene2GO = geneID2GO, nodeSize = 10)
GOdataCC_X <- new("topGOdata", ontology = c("CC"), allGenes = geneList,
  geneSel = MyInterestingGenes_X, annot = annFUN.gene2GO, gene2GO = geneID2GO, nodeSize = 10)
```

### results - default 
Using topGO weight/elim combo algorithm, fisher stat
```{r} 
# define test using the weight01 algorithm (default) with fisher
resultFisher_BP_X<- runTest(GOdataBP_X, algorithm = "weight01",statistic = "fisher")
resultFisher_MF_X<- runTest(GOdataMF_X, algorithm = "weight01",statistic = "fisher")
resultFisher_CC_X<- runTest(GOdataCC_X, algorithm = "weight01",statistic = "fisher")
```
### results - default
note: p-values need FDR correction
```{r} 
allRes_BP <- GenTable(GOdataBP_X, ranksOf = "classic", orderBy = "weightFisher",
                        topNodes = 50,classic = resultFisher_BP_X)
allRes_MF <- GenTable(GOdataMF_X, classic = resultFisher_MF_X, orderBy = "weightFisher", 
                      ranksOf = "classic", topNodes = 50)
allRes_CC <- GenTable(GOdataCC_X, classic = resultFisher_CC_X, orderBy = "weightFisher", 
                      ranksOf = "classic", topNodes = 50)
#showSigOfNodes(GOdataBP_X, score(allRes_BP$classic), firstSigNodes = 5, useInfo = 'all')
#showSigOfNodes(GOdataBP_X,termsP.value = allRes_BP$classic, = 5)

print(allRes_BP)
print(allRes_MF)
print(allRes_CC)
```

```{r} 
allGO_BP=usedGO(GOdataBP_X)
#performing BH correction on our p values
p.adj_BP=round(p.adjust(allRes_BP$classic,method="BH"),digits = 4)
 
# create the file with all the statistics from GO analysis
allResBH_BP=cbind(allRes_BP,p.adj_BP)
allResBH_BP=allResBH_BP[order(allResBH_BP$p.adj_BP),]
 
#get list of significant GO before multiple testing correction
results.table.p.bp= allResBH_BP[which(allResBH_BP$classic<=0.05),]
print(results.table.p.bp) 
#get list of significant GO after multiple testing correction
results.table.bh.bp=allResBH_BP[which(allResBH_BP$p.adj<=0.1),]
print(results.table.bh.bp)
# just two top hits
# GO:0006355 regulation of DNA-templated transcriptio...	
# GO:0016192 vesicle-mediated transport	
#####
#save first top 50 ontolgies sorted by adjusted pvalues
#write.table(all_res_final[1:50,],"summary_topGO_analysis.csv",sep=",",quote=FALSE,row.names=FALSE)
 
# PLOT the GO hierarchy plot: the enriched GO terms are colored in yellow/red according to significance level
 
# pdf(file='topGOPlot_BP_fullnames.pdf', height=12, width=12)
# showSigOfNodes(GOdataBP_X, score(resultFisher_BP_X), useInfo = "all", sigForAll=FALSE, firstSigNodes=2,.NO.CHAR=50)
# dev.off()
# 
# all_res_CC=GenTable(GOdataCC_X, weightFisher=weight_fisher_result_CC, orderBy='weightFisher', topNodes=length(GOdataCC_X))
# showSigOfNodes(GOdataCC_X, score(weight_fisher_result_CC), useInfo = "all", sigForAll=FALSE, firstSigNodes=2,.NO.CHAR=50)
# 
# showSigOfNodes(GOdataMF_X, score(weight_fisher_result_MF), useInfo = "all", sigForAll=FALSE, firstSigNodes=2,.NO.CHAR=50)
```

### Redo again with background of just X genes
```{r}
GO_map_denom_x <- GO_map_denom %>% filter(ID %in% GO_anno_X$anno)
write.table(GO_map_denom_x,"geneIDsGO_denom.map", row.names=F,col.names = F,quote = F)

geneID2GO_x<-readMappings(file="geneIDsGO_denom.map",sep=" ",IDsep=",") 

geneNames_x <- names(geneID2GO_x)

MyInterestingGenes_X<-as.character(Xgenelist$hap1)
geneList_x <- factor(as.integer(geneNames_x %in% MyInterestingGenes_X))
names(geneList_x) <- geneNames_x
str(geneList_x) # 1649 total genes (Vs 5000 ish from before)
GOdataBP_Xonly <- new("topGOdata", ontology = c("BP"), allGenes = geneList_x,description = "X genes in total set",
  geneSel = MyInterestingGenes_X, annot = annFUN.gene2GO, gene2GO = geneID2GO_x, nodeSize = 10)
GOdataMF_Xonly <- new("topGOdata", ontology = c("MF"),
                      allGenes = geneList_x,
                      description = "X genes in total set",
                      geneSel = MyInterestingGenes_X, 
                      annot = annFUN.gene2GO, gene2GO = geneID2GO_x, nodeSize = 10)
GOdataCC_Xonly <- new("topGOdata", ontology = c("CC"),  allGenes = geneList_x,
                      description = "X genes in total set",
                      geneSel = MyInterestingGenes_X, 
                      annot = annFUN.gene2GO, gene2GO = geneID2GO_x, nodeSize = 10)
```


```{r}
# define test using the weight01 algorithm (default) with fisher
resultFisher_BP_Xonly<- runTest(GOdataBP_Xonly, algorithm = "weight01",statistic = "fisher")
resultFisher_MF_Xonly<- runTest(GOdataMF_Xonly, algorithm = "weight01",statistic = "fisher")
resultFisher_CC_Xonly<- runTest(GOdataCC_Xonly, algorithm = "weight01",statistic = "fisher")
```
### results - default
note: p-values need FDR correction
```{r} 
allRes_BPonly <- GenTable(GOdataBP_Xonly, ranksOf = "classic", orderBy = "weightFisher",
                        topNodes = 50,classic = resultFisher_BP_Xonly)
allRes_MFonly <- GenTable(GOdataMF_Xonly, classic = resultFisher_MF_Xonly, orderBy = "weightFisher", 
                      ranksOf = "classic", topNodes = 50)
allRes_CConly <- GenTable(GOdataCC_Xonly, classic = resultFisher_CC_Xonly, orderBy = "weightFisher", 
                      ranksOf = "classic", topNodes = 20)
#showSigOfNodes(GOdataBP_X, score(allRes_BP$classic), firstSigNodes = 5, useInfo = 'all')
#showSigOfNodes(GOdataBP_X,termsP.value = allRes_BP$classic, = 5)
print(allRes_BPonly)
print(allRes_MFonly)
print(allRes_CConly)

allGO_BPx=usedGO(GOdataBP_Xonly)
#performing BH correction on our p values
p.adj_BP=round(p.adjust(allRes_BPonly$classic,method="BH"),digits = 4)
 
# create the file with all the statistics from GO analysis
allResBH_BPonly=cbind(allRes_BPonly,p.adj_BP)
allResBH_BPonly=allResBH_BPonly[order(allResBH_BPonly$p.adj_BP),]
 
#get list of significant GO before multiple testing correction
results.table.p.bp.x= allResBH_BPonly[which(allResBH_BPonly$classic<=0.05),]
print(results.table.p.bp.x) 
#get list of significant GO after multiple testing correction
results.table.bh.bp.x=allResBH_BPonly[which(allResBH_BPonly$p.adj<=0.1),]
print(results.table.bh.bp.x) 

### nope

showSigOfNodes(GOdataBP_Xonly, score(resultFisher_BP_Xonly), useInfo = "all", sigForAll=FALSE, firstSigNodes=2,.NO.CHAR=50)

## try MF and CC 

allGO_MFx=usedGO(GOdataMF_Xonly)
#performing BH correction on our p values
p.adj_MF=round(p.adjust(allRes_MFonly$classic,method="BH"),digits = 4)
 
# create the file with all the statistics from GO analysis
allResBH_MFonly=cbind(allRes_MFonly,p.adj_MF)
allResBH_MFonly=allResBH_MFonly[order(allResBH_MFonly$p.adj_MF),]
 
#get list of significant GO before multiple testing correction
results.table.p.mf.x= allResBH_MFonly[which(allResBH_MFonly$classic<=0.05),]
print(results.table.p.mf.x) 
#get list of significant GO after multiple testing correction
results.table.bh.mf.x=allResBH_MFonly[which(allResBH_MFonly$p.adj<=0.1),]
print(results.table.bh.mf.x) 

### nope

showSigOfNodes(GOdataBP_Xonly, score(resultFisher_BP_Xonly), useInfo = "all", sigForAll=FALSE, firstSigNodes=2,.NO.CHAR=50)
##
results.table.p.cc.x= allRes_CConly[which(allRes_MFonly$classic<=0.05),]
print(results.table.p.cc.x)
## womp!
myterms =results.table.p.bp.x$GO.ID # change it to results.table.bh$GO.ID if working with BH corrected values
mygenes = genesInTerm(GOdataBP_Xonly, myterms)
 
var=c()
for (i in 1:length(myterms))
{
myterm=myterms[i]
mygenesforterm= mygenes[myterm][[1]]
mygenesforterm=paste(mygenesforterm, collapse=',')
var[i]=paste("GOTerm",myterm,"genes-",mygenesforterm)
}
var
#write.table(var,"genetoGOmapping.txt",sep="\t",quote=F)
```
Do it over again but with the retained x genes... y'all
```{r}
## hrmgpslgjklsfsg
#GO_map_denom_x <- GO_map_denom %>% filter(ID %in% GO_anno_X$anno)
#write.table(GO_map_denom_x,"geneIDsGO_denom.map", row.names=F,col.names = F,quote = F)

geneID2GO_x<-readMappings(file="geneIDsGO_denom.map",sep=" ",IDsep=",") 

geneNames_x <- names(geneID2GO_x)
retainedgenelist<-GO_map_denom_x %>% filter(!ID %in% MyInterestingGenes_X)

MyInterestingGenes_retained<-as.character(retainedgenelist$ID)
geneList_retained <- factor(as.integer(geneNames_x %in% MyInterestingGenes_retained))
names(geneList_retained) <- geneNames_x
str(geneList_retained) # 1649 total genes (Vs 5000 ish from before)
GOdataBP_ret <- new("topGOdata", ontology = c("BP"), allGenes = geneList_retained,description = "X genes in total set",
  geneSel = MyInterestingGenes_retained, annot = annFUN.gene2GO, gene2GO = geneID2GO_x, nodeSize = 10)
GOdataMF_Xonly <- new("topGOdata", ontology = c("MF"),
                      allGenes = geneList_x,
                      description = "X genes in total set",
                      geneSel = MyInterestingGenes_X, 
                      annot = annFUN.gene2GO, gene2GO = geneID2GO_x, nodeSize = 10)
GOdataCC_Xonly <- new("topGOdata", ontology = c("CC"),  allGenes = geneList_x,
                      description = "X genes in total set",
                      geneSel = MyInterestingGenes_X, 
                      annot = annFUN.gene2GO, gene2GO = geneID2GO_x, nodeSize = 10)

#GOdataBP_ret
allGO = usedGO(object = GOdataBP_ret) 
# use it in GenTable as follows:
GenTable(GOdata, ... ,topNodes = length(allGO))
resultFisher_BP_ret<- runTest(GOdataBP_ret, algorithm = "weight01",statistic = "fisher")
resultElim_BP_ret<- runTest(GOdataBP_ret, algorithm = "elim",statistic = "fisher")
resultElimKS_BP_ret<- runTest(GOdataBP_ret, algorithm = "elim",statistic = "KS")


allRes_BP_ret <- GenTable(GOdataBP_ret, ranksOf = "classicFisher", orderBy = "classicFisher",
                        topNodes = length(allGO),
                        classicFisher = resultFisher_BP_ret,
                        elimFisher=resultElim_BP_ret,elimKS=resultElimKS_BP_ret)
#head(score(resultFisher_BP_ret))
#pvalFis <- score(resultFisher_BP_ret)
#head(pvalFis)
#hist(pvalFis, 50, xlab = "p-values")

```

